---
title: Introduction to R
author:  Alistair Bailey and Data Carpentry contributors
minutes: 45
---


```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Assign values to objects and variables in R.
> * Define the following terms as they relate to R: script, function, working directory, assign, object, variable.
> * Describe what vectors are and how they are manipulated in R.
> * Inspect the content of vectors in R and manipulate their content.
> * Extract elements from vectors in R.
> * Employ logic (i.e. TRUE, FALSE) to subset data in a vector.
> * Analyze vectors with missing data.
> * Call functions with zero or more named or unnamed arguments.

------------

The aim is get us working with, and exploring the data as soon as possible,
but first we will step back to cover some R fundamentals.

## Entering input, assigning objects and evaluation

We'll start in the console. At the prompt we type expressions using the
`<- ` assignment operator to create objects. In the example below we create 
an object called `weight` with a value of `55`. 

To evaluate the expression, that is to create `weight`, we simply press 
<kbd>Enter</kbd>. 

(Tips: pushing <kbd>Alt</kbd> at the same time as the
<kbd>-</kbd> key) will write ` <- `)


```{r, purl=FALSE}
# Assign weight a value of 55
weight <- 55
```

Having created `weight`, evaluating the object itself by typing it's name,
will print the object to the console:

```{r, purl=FALSE, results='show'}
# Assign weight a value of 55
weight <- 55
# Print the value of the object weight
weight
```

## R objects

R has five basic (atomic) classes of objects:

+ character
+ numeric (real numbers)
+ integers (whole numbers)
+ complex numbers
+ logical (True/False)

Collections of objects are vectors, and vectors will only store one class of 
basic object. If you mix classes e.g. characters and integers, R will coerce
them to be all the same class e.g. your integers may become characters.

## Vectors

### Inspecting and extracting elements from vectors

### Subsetting vectors

## Functions

### Function help files

## Dealing with missing data

