---
title: Manipulating and analyzing data with dplyr; Exporting data
author: Data Carpentry contributors
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
surveys <- read.csv("data/portal_data_joined.csv")
```

```{r, echo=FALSE, purl=TRUE}
## The dplyr lesson
```

------------

> ### Learning Objectives
>
> By the end of this lesson the learner will:
>
> * Recall the six data manipulation ‘verbs’ in the dplyr package and know what they do.
> * Select subsets of columns and filter rows in a data.frame according to a condition(s).
> * Employ the ‘pipe’ operator to link together a sequence of dplyr commands.
> * Employ the ‘mutate’ command to apply functions to existing columns and create new columns of data.
> * Export a data.frame to a .csv file.

------------

# Data Manipulation using dplyr

## Motivation

In this lesson we'll transform the data using `dplyr` to:

+ change the focus. For example, filtering only a few species of rodent or 
a particular time period.
+ create new variables that are functions of existing variables. 
+ calculate summary statistics of the rodent data, such as mean weight.
+ create a tidy dataset of complete sets of observations.

## What is tidy data?

## `dplyr` verbs

## Piping operations

## Split-apply-combine

